---
title: 'CSU Coding and Cookies: Data Visualization Using ggplot2'
output: html_document
---

```{r setup, include=FALSE, message = FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Un-comment the following line if tidyverse is not installed on your machine.
# install.packages("tidyverse")
library( "tidyverse" )
```

Based on: <https://datacarpentry.github.io/R-ecology-lesson/visualizing-ggplot.html>

# Welcome and Introductions

Welcome

Instructor(s) introduction(s)

Participants: Tell us

- your name,
- department and
- Do you have any experience with data visualization? If so, what kinds of graphs have you made?

# Outline

- Why `ggplot`?
- How to use it
  - basic elements (data, aesthetics, geoms)
  - modifications (transparency, color, grouping)
  - themes (modifying default, using premade)
- Exporting plots (`ggsave`)

Today we will learn how to use the `ggplot2` package to make (beautiful!) graphs. 

`ggplot2` is part of `tidyverse` which also includes `dplyr`, `tidyr` and more.

# Why `ggplot2`?

- One of the most popular data visualization packages
- Makes highly customizable graphs, suitable for publication
- Layered syntax via `+` operator
- Automatic Legends and Scaling
- Part of the `tidyverse` ecosystem
- General help page: <https://ggplot2.tidyverse.org>

# Getting Started 

## Organize your files

1.  Create a folder called `CandC_R_ggplot` somewhere on your computer.
2.  Download workshop files and save them in your `CandC_R_gglot` folder (that you just created).

<https://github.com/CSU-Stat-Lab/Coding-and-Cookies/>

- `CandC_R_ggplot.Rmd`
- `surveys_complete_77_89.csv`

Keeping the `.Rmd` and `.csv` files in the same exact folder will make importing data easier.

3.  Open `CandC_R_Basics.Rmd` in RStudio.

## Load Packages

`tidyverse` is loaded in the first code chunk of this `.Rmd` file.  Today, we will not require any further packages.

## Data Introduction

**Data set**: A survey of small animals.  

Data are stored as a data frame object in R.

**Instructor**: show them where the data are to download and to where they should be downloaded.


- **Rows**: observations of individual animals
- **Columns**: Variables that describe the animals

Because the `.Rmd` and `.csv` files are in the same folder, we can read the file using:

```{r}
## Code obtained from Data Carpentry Ecology Lesson
####  http://www.datacarpentry.org/R-ecology-lesson/04-visualization-ggplot2.html

# Load the data 
surveys_complete <- read.csv("surveys_complete_77_89.csv")
```

- Take a look at the data

```{r}
str( surveys_complete )
```

## Check In

**Instructor:**

- From Packages tab, show `ggplot` loaded (checked) 
- Show `surveys_complete` in the environment.
- Check if anyone had trouble importing the data or loading ggplot2. 


- It is generally a good idea to convert character variables to type factor if they are actually factor variables.
    - Much of the code below will work as expected if they are kept as character variables, but it is a good habit to convert them as some functions will treat character variables differently from factor variables.

- Recall that the "pipe" operator in base R is the combination of symbols `|>`, which can be read as the (adverb) "and then".  

In the code below, the first line assigns `surveys_complete` to itself (if you don't do this step, it will not save what comes next after the `|>` instances), "and then" using the `surveys_complete` object, it converts `species_id` to a factor, and so on.  Note that multiple pipe commands are employed.

```{r}
surveys_complete <- surveys_complete |>
  mutate( species_id = as_factor( species_id ) ) |>
  mutate( sex = as_factor( sex ) ) |>
  mutate( genus = as_factor( genus ) ) |>
  mutate( species = as_factor( species ) ) |>
  mutate( taxa = as_factor( taxa ) ) |>
  mutate( plot_type = as_factor( plot_type ) )

str( surveys_complete )
```

# Opening Examples

**Instructor**: before spending time on all the details, demonstrate what `ggplot2` can do with the following examples.  

## Example 1

```{r}
ggplot(data = surveys_complete, 
            aes(x = weight, 
                   y = hindfoot_length,
    color = genus,
    group = genus)) + 
geom_point(alpha = 0.2) + 
geom_smooth() +
labs(x = "weight (g)", y = "hindfoot length (mm)",title = "Example 1")

```

In just a couple of lines, we asked `ggplot2` to:

-	take a data set (surveys_complete)
-	create a scatterplot (geom_point) of hindfoot_length vs weight (aes)
-	color and group by genus with legend created automatically
-	add a smooth curve by genus (geom_smooth)
-	customize the axis labels and title (labs)

We don't need to understand every part of this code yet. By the end of the workshop, you will understand what these pieces mean.

## Example 2

```{r}
ggplot(data = surveys_complete, 
            aes(x = weight, 
                   y = hindfoot_length)) + 
geom_point(alpha = 0.2) + 
geom_smooth() +
facet_wrap( ~ genus) +
labs(x = "weight (g)",
        y = "hindfoot length (mm)",
        title = "Example 2")
```

Wow! We made 8 neatly formatted and labeled plots with minimal effort.  


# Scatterplots

## Basic graph elements

`ggplot(data = <DATA>,` `mapping = aes(<MAPPINGS>)) +` `<GEOM_FUNCTION>()`

- `ggplot()` creates a plot. Must give it:
  - a data object `ggplot( data = surveys_complete )` 
  - a mapping using `aes`
  - then add layers, such as what type of plot to make. Many plot types are available, such as
    - `geom_point` (a point, or scatter, plot)
    - `geom_jitter` (a point plot whose points are jittered to limit overlapping points)
    - `geom_dotplot` (similar to `geom_jitter` but cleaner)
    - `geom_boxplot` (box plots)
    - `geom_bar` (bar plots: creates data counts before plotting)
    - `geom_col` (like `geom_bar` but plots the actual data without first making counts)
    - `geom_histogram`

- **Step 1**: specify the data to be plotted. The following code opens a plot window, but does not graph anything

The following code simply makes a blank plot.

```{r}
ggplot( data = surveys_complete )
```


- **Step 2**: specify what to put on each axis. The following code opens a plot window and draws axes.

The following makes a blank plot but with axes added.

```{r}
ggplot(data = surveys_complete, mapping = aes(x = weight, 
                                      y = hindfoot_length))
```

- **Step 3**: specify the geometry. opens a plot window, draws axes, and draw points for each row of the data frame.

```{r}
ggplot(data = surveys_complete, 
       aes(x = weight, y = hindfoot_length)) +
  geom_point()
```

- **Alternate Step 3**: plotting with `dplyr`.

```{r}
surveys_complete |>
  ggplot(aes(x = weight, y = hindfoot_length)) +
  geom_point()
```

- Add transparency with the alpha argument to `geom_point`.

```{r}
ggplot(data = surveys_complete, 
       aes(x = weight, y = hindfoot_length)) + 
  geom_point(alpha = 0.1)
```

- Add color by species with color argument to `aes`.

```{r}
ggplot(data = surveys_complete, 
       aes(x = weight, y = hindfoot_length, color = species_id)) + 
  geom_point(alpha = 0.1)
```

- `aes` argument to `ggplot` specifies for the whole graph.

```{r}
ggplot(data = surveys_complete, 
       aes(x = weight, y = hindfoot_length, color=species_id)) + 
  geom_point(alpha = 0.1)
```

### Your turn

Use the previous example as a starting point.

- Add color to the data points according to the plot from which the sample was taken (`plot_id`).

**Hint**: Check the class for `plot_id`. Consider changing the class of `plot_id` from integer to factor. Why does this change how R makes the graph?

```{r}

```

# Boxplots

- Make a boxplot

```{r}
ggplot(data = surveys_complete, 
       aes(x = species_id, y = hindfoot_length)) +
  geom_boxplot()
```

**Instructor:** briefly remind everyone what a boxplot shows (ex: median, min, max)

- Overlay points on a boxplot

```{r}
ggplot(data = surveys_complete, 
       aes(x = species_id, y = hindfoot_length)) + 
  geom_boxplot() +
  geom_jitter()
```

**Instructor**: Explain what is meant by "jitter".

## Your turn

Make the same plot as above, but use `geom_point` instead of `geom_jitter`.

### Your turn

Plot the same data as in the previous example, but as a Violin plot

**Hint**: see `?geom_violin`.

- What information does this plot give you about the data that a box plot does not?

```{r}

```

# Time series data

- reshape the data

**Instructor**: show the features of `tally` and explain what happens here.

```{r}
yearly_counts <- surveys_complete |> 
  group_by(year, species_id) |> 
  tally()
```

- Plot `n` vs. `year`

```{r}
ggplot(data = yearly_counts, aes(x = year, y = n)) + 
  geom_line()
```

- One line for each species

```{r}
ggplot(data = yearly_counts, 
       aes(x = year, y = n, group = species_id)) + 
  geom_line()
```

- add color by species

```{r}
ggplot(data = yearly_counts, 
       aes(x = year, y = n, group = species_id, 
           color = species_id)) +
  geom_line()
```

### Your turn

Use what you just learned to create a plot that depicts how the average weight of each species changes through the years.

**Hint** the functions `group_by` along with `summarize` may be very useful here.  It might also be useful to remove missing values before calling these functions.

```{r}

```

# Themes

## Using pre-made themes

- Apply a theme (see `?theme_bw` for other themes)

```{r}
ggplot(data = yearly_counts, 
       aes(x = year, y = n, color = species_id)) + 
  geom_line() + 
  theme_bw() 
```

## Customizing themes

- Change axis labels and titles

```{r}
ggplot(data = yearly_counts, 
       aes(x = year, y = n, color = species_id)) + 
  geom_line() + 
  labs(title = 'Observed species in Time', 
       x = 'Year of observation', 
       y = 'Count', color = 'Species') + 
  theme_bw()
```

# Layering geoms

It is possible to layer more than one type of plot onto another (e.g., when we added points to our box plot example).  Here is another example where we add a smooth curve fit to a point plot.

```{r}
ggplot(data = surveys_complete, 
            aes(x = weight, 
                   y = hindfoot_length,
    color = genus,
    group = genus)) + 
geom_point(alpha = 0.2) + 
geom_smooth() +
labs(x = "weight (g)", y = "hindfoot length (mm)",title = "Example 1")

```

# Creating panels of plots

Panels of plots are called "facets" in `ggplot2`.  Here is an example using `facet_wrap`.

**Instructor**: Describe the formula argument.

```{r}
ggplot(data = surveys_complete, 
            aes(x = weight, 
                   y = hindfoot_length)) + 
geom_point(alpha = 0.2) + 
geom_smooth() +
facet_wrap( ~ genus) +
labs(x = "weight (g)",
        y = "hindfoot length (mm)",
        title = "Example 2")
```

# Save your plot

## save plot to a variable

```{r}
my_plot <- ggplot(data = yearly_counts, 
                  aes(x = year, y = n, color = species_id)) + 
  geom_line() + 
  labs(title = 'Observed species in time', 
       x = 'Year of observation', 
       y = 'Number of species', color = 'Species') + 
  theme_bw() + 
  theme(axis.text.x = element_text(colour="grey20", 
                                   size=12, angle=90, 
                                   hjust=.5, vjust=.5), 
        axis.text.y = element_text(colour="grey20", size=12), 
        text=element_text(size=16, family="Arial"))
```

## save plot to a file

```{r}
ggsave("name_of_file.png", my_plot, width=15, height=10)
```

Finally, it is good practice to cite software and packages that you use for your publications.  To cite `ggplot2` (and similarly for other packages) run the following code:

```{r}
citation("ggplot2")
```

# Need Help?

Learning R is an ongoing process. You do not need to memorize everything.

Useful resources include:

- CSU Statistics Laboratory (<https://statlab.colostate.edu/>)
- R documentation and help pages
- R cheat sheets (<https://opensource.posit.co/resources/cheatsheets/>)
- CSU Libraries Coding Support Hub (<https://codingsupport.colostate.edu/r/>)
- CSU Data Management Services (<https://lib.colostate.edu/services/data-management/>)

