---
title: "CSU Coding and Cookies: Tidy Data in R"
output: html_document
---

```{r setup, include=FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo=TRUE)

# install.packages("pacman")
pacman::p_load(knitr, tidyverse, fivethirtyeight)
```

# Welcome and Introductions

Welcome!

Instructor(s) introduction(s)

Participants: Tell us

- your name,
- department, and
- Do you have any experience with data cleaning or reformatting? If so, what kinds of tasks have you done?

# Outline

Today we will learn how to use the `tidyverse` package to manipulate data.

The `tidyverse` package is a collection of packages that includes: `dplyr`, `tidyr`, `ggplot2`, and many more.

Topics include

- filtering rows
- selecting columns
- creating new variables
- grouping data
- summarizing data
- reshaping data

The goal is to learn a small number of functions that can accomplish many common data management tasks.

# Getting Started

## Organize your files

1.  Create a folder called `CandC_R_Tidy_Data` somewhere on your computer.
2.  Download workshop files and save them in your `CandC_R_Tidy_Data` folder (that you just created).

<https://csu-stat-lab.github.io/Coding-and-Cookies/>

- `CandC_R_Tidy_Data.Rmd`

3.  Open `CandC_R_Tidy_Data.Rmd` in RStudio.

## Load Packages

`tidyverse` is loaded in the first code chunk of this `.Rmd` file.  Today, we will not require any further packages.

## Load the data

Today we will use the **biopics** data from the **fivethirtyeight** package.

```{r}
citation("fivethirtyeight")

data("biopics")

```

Before working with any data set, it is a good idea to understand what it contains.

```{r}
str(biopics)
head(biopics)
summary(biopics)
```

Notice that

- each row represents one movie
- each column represents one variable
- different columns have different data types

### Your Turn

Answer the following questions.

- What are these data about (hint: see the help file for the data set)? 
- How many observations are in the data set?
- How many variables?
- Which variables are numeric?
- Which variables are character?

# The dplyr verbs

Most data manipulation in the tidyverse is done with a handful of functions, often called **verbs**.

Some of the most useful are

- `filter()`
- `select()`
- `mutate()`
- `group_by()`
- `summarize()`
- `arrange()`

These functions all take a data frame (or tibble) as input and return another tibble.

A `data.frame` object is the primary data structure from base R.  A `tibble` object is very similar, but is part of `tidyverse`.

# Filtering Rows

Often we want only observations meeting certain conditions.

The function `filter()` keeps rows that satisfy one or more logical conditions.

## Example

Select only films released in the UK after 1990.

```{r}
biopicsUK1990 <-
    filter(biopics,
           country=="UK",
           year_release>1990)

biopicsUK1990
```

Notice that

- the first argument is always the data frame
- additional arguments are logical conditions
- only rows meeting **all** conditions are kept

Common comparison operators include

|Operator|Meaning|
|---------|-------|
|`==`|equal to|
|`!=`|not equal to|
|`>`|greater than|
|`<`|less than|
|`>=`|greater than or equal|
|`<=`|less than or equal|

### Your Turn

Create a data frame called `biopCan` containing only movies from Canada.

```{r}

```

### Your Turn

Create a data frame containing only movies released after 2010.

```{r}

```

# Selecting Columns

Sometimes we want fewer variables rather than fewer observations.

The `select()` function keeps only the columns we specify.

```{r}
smallData <-
    select(biopics,
           title,
           year_release,
           country)

smallData
```

Notice that the number of rows stays the same, but the number of columns changes.

### Your Turn

Create a data frame containing only

- title
- lead_actor_actress
- box_office

```{r}

```

# Combining Operations with the Pipe

Instead of creating many intermediate data sets, we can connect operations together using the pipe operator `|>`.

Compared to the `dplyr` “verbs”, the pipe operator acts like the word (adverb) “then”.

The pipe sends the output from one step into the next step.

Instead of

```r
x <- filter(...)
y <- select(x,...)
```

we can write

```{r}
biopics |>
    filter(country=="UK",
           year_release>1990) |>
    select(title,
           subject_sex,
           box_office)
```

Many people find this easier to read because the analysis proceeds from top to bottom.

### Your Turn

Write a pipeline that

- keeps only Canadian movies
- selects only the title and box office columns

```{r}

```

# Creating New Variables

The `mutate()` function creates new columns.

Suppose we want to know which decade each movie belongs to.

```{r}
biopicsDec <-
    biopics |>
    mutate(decade=floor(year_release/10)*10)

head(biopicsDec)
```

The original variables remain, and a new column is added.

### Your Turn

Create a variable called `box_millions` that equals `box_office/1000000`

```{r}

```

# Grouping Data

Many analyses involve calculating summaries for groups.

The `group_by()` function tells R which groups should be analyzed separately.

By itself, `group_by()` changes nothing that you can see.

```{r}
biopics |>
    group_by(country)
```

Its usefulness comes when it is combined with `summarize()`.

# Summarizing Data

The `summarize()` function reduces many observations to one or more summary statistics.

For example, compute the mean box office by country.

```{r}
biopics |>
    drop_na(box_office) |> # See ?drop_na to learn about this function.
    group_by(country) |>
    summarize(
        Mean=mean(box_office),
        SD=sd(box_office),
        N=n()
    )
```

Notice

- missing values were removed first
- each country now has one row
- we calculated three summaries at once

### Your Turn

Calculate

- the mean box office
- the sample size

for each value of `subject_sex`.

```{r}

```

# Sorting Results

The `arrange()` function changes the order of rows.

```{r}
biopics |>
    drop_na(box_office) |>
    group_by(country) |>
    summarize(
        Mean=mean(box_office)
    ) |>
    arrange(desc(Mean))
```

The country with the largest average box office now appears first.

### Your Turn

Arrange the countries from smallest to largest average box office.

```{r}

```

# Reshaping Data

Sometimes we want data to be wider or longer.

The **tidyr** package provides two important functions

- `pivot_wider()`
- `pivot_longer()`

## Wide Data

Create one column for each subject sex.

```{r}
dec_box <- biopicsDec |>
    drop_na(box_office) |>
    group_by(decade, subject_sex) |>
    summarize(Mean=mean(box_office)) |>
    pivot_wider(
        names_from=subject_sex,
        values_from=Mean
    )
```

## Long Data

Convert summary statistics into a longer format.

```{r}
dec_box |>
    pivot_longer(
        cols=-decade,
        names_to="Statistic",
        values_to="Value"
    )
```

### Your Turn

Look at the output of each pivot.

Discuss with someone near you

- Which output is wider?
- Which output is longer?
- When might each format be useful?

# Putting Everything Together

The following pipeline performs a complete analysis.

```{r}
biopics |>
    drop_na(box_office) |>
    mutate(decade=floor(year_release/10)*10) |>
    group_by(decade) |>
    summarize(
        Mean=mean(box_office),
        SD=sd(box_office),
        N=n()
    ) |>
    arrange(decade)
```

Notice how each step performs one simple task.

This style of programming is one of the major strengths of the tidyverse.

# Need Help?

Learning R is an ongoing process. You do not need to memorize everything.

Useful resources include:

- CSU Statistics Laboratory (<https://statlab.colostate.edu/>)
- R documentation and help pages
- R cheat sheets (<https://opensource.posit.co/resources/cheatsheets/>)
- CSU Libraries Coding Support Hub (<https://codingsupport.colostate.edu/r/>)
- CSU Data Management Services (<https://lib.colostate.edu/services/data-management/>)

